---
title: "HW 3"
author: "Steven Brannum"
date: "`r Sys.Date()`"
output: html_document
---

# Part A

```{r packages}
library(Matrix)
library(arules)
library(arulesViz)
```

```{r}
# crash <-read.csv("C:/Users/User/SkyDrive/SP18/Applied Data Mining - BIA 6301 BSB/HW3/HW3/HW3/crash.csv")
crash <-read.csv("D:/OneDrive/SP18/Applied Data Mining - BIA 6301 BSB/HW3/HW3/HW3/crash.csv")
#crash <- read.transactions("D:/OneDrive/SP18/Applied Data Mining - BIA 6301 BSB/HW3/HW3/HW3/crash.csv")

dim(crash)
head(crash)

```

Regions categorization

```{r}
labels.midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota")
labels.northeast <- c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania")
labels.south <- c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "District of Columbia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas")
labels.west <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington")


crash$Region <- crash$State
levels(crash$Region) <- list('midwest' = labels.midwest, 'northeast' = labels.northeast, 'south' = labels.south, 'west' = labels.west)

table(crash$Region)

```


Drop columns (state, atmospheric condition, crash.date, fatalities, person.type, race, crash.date.year, crash date.day-of-month, crash.date.day.of.week)

```{r}
crash <- crash[, - c(1, 2, 3, 4, 8, 10, 13, 15, 16)]
head(crash)
```

Standardize drug involvement to Unknown, No, and Yes

```{r}
crash$Drug.Involvement <- factor(ifelse(crash$Drug.Involvement == '\\N', 'Unknown',
                                    ifelse(crash$Drug.Involvement == 'Not Reported', 'Unknown',
                                        ifelse(crash$Drug.Involvement == 'Yes', 'Yes', 'No'))))

table(crash$Drug.Involvement)

```

Categorize Alcohol Results to AboveLegal, Legal, or Unknown

```{r}

crash$Alcohol.Results <- sapply(crash$Alcohol.Results, function(y) { if (is.na(y)) 'Unknown' else if (y < .09) 'Legal' else 'AboveLegal' })

crash$Alcohol.Results <- factor(crash$Alcohol.Results)
table(crash$Alcohol.Results)

```


Categorize age

```{r}

crash$Age <- factor(ifelse(crash$Age < 15, "Kid",
                   ifelse(crash$Age < 30, "YoungAdult",
                           ifelse(crash$Age < 45, "Adult",
                                   ifelse(crash$Age < 60, "OlderAdult", "Retiring")))))

table(crash$Age)
```


Turn roadways into rural or urban and change NA to unknown because

```{r}
#replace hyphens with space
crash$Roadway <- gsub('-', ' ', crash$Roadway)
head(crash)

#remove everything after first space
crash$Roadway <- gsub(' [A-z ]*', '', crash$Roadway)
head(crash)

#change \\N to unknown
crash$Roadway <- sapply(crash$Roadway, function(y) { if (y == '\\N') 'Unknown' else y })

table(crash$Roadway)

#factor
crash$Roadway <- factor(crash$Roadway)

```

Factor Crash Month

```{r}
crash$Crash.Date.month <- factor(crash$Crash.Date.month, labels = c("Jan", "Feb", "Mar", "April", "May", "June", "July", "Aug", "Sept", "Oct", "Nov", "Dec"))
table(crash$Crash.Date.month)

```



Frequencies

```{r}
# create transactions
head(crash)
tcrash <- as(crash, "transactions")
itemFrequencyPlot(tcrash, topN = 20)

basic_rules <- apriori(tcrash, parameter = list(support = 0.1, confidence = 0.5, minlen = 2))
print(basic_rules)
inspect(basic_rules[1:10])

#remove redundant
basic_rules_pruned <- basic_rules[!is.redundant(basic_rules)]
print(basic_rules_pruned)
inspect(basic_rules_pruned[1:10])

```

Remove unknown drug and alcohol results from the rhs because they're irrelevant


```{r}
#basic_rules_rhsunknown_removed <- subset(basic_rules_pruned, !(rhs %in% 'Alcohol.Results=Unknown'))
#basic_rules_rhsunknown_removed <- subset(basic_rules_rhsunknown_removed, !(rhs %in% 'Drug.Involvement=Unknown'))

#basic_rules_rhsunknown_removed <- subset(basic_rules_rhsunknown_removed, (rhs %pin% 'Gender'))

#sort
basic_rules_sorted <- sort(basic_rules_pruned, by = c("lift", "confidence"))
inspect(basic_rules_sorted[1:15])

table(crash$Gender)
```




# Part B

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(ROCR)
library(knitr)
library(e1071)
library(class)
```

```{r}
bank.df <- read.csv("D:/OneDrive/SP18/Applied Data Mining - BIA 6301 BSB/HW1/HW1/HW1/UniversalBank.csv")
bank.df <- bank.df[,-c(1,5)] # drop ID and zip code columns.  
# create categorical variable for education
bank.df$Education <- factor(bank.df$Education, levels = c(1, 2, 3), labels = c("Undergrad", "Graduate", "Advanced/Professional"))

bank.df$Personal.Loan <- factor(bank.df$Personal.Loan, levels = c(0, 1), labels = c("NotInterested", "Interested"))


bankData <- bank.df[, c(8, 1:7, 9:12)]
bankData <- bankData[1:5000,]

trainIndex <- createDataPartition(bankData$Personal.Loan, p = .8, list = FALSE, times = 1)
trainSet <- bankData[trainIndex,]
testSet <- bankData[-trainIndex,]
dim(trainSet)
dim(testSet)

head(bank.df)


```


```{r nbcompare}
options(warn=-1)

cvCtrl <- trainControl(method = "cv", number = 10)
set.seed(312)

actual <- testSet$Personal.Loan

#nb
head(testSet)
trainSetNB <- trainSet[-c(7)]
testSetNB <- testSet[-c(7)]
bank_cv_nb <- train(Personal.Loan ~ ., data = trainSet, method = "naive_bayes", metric = "Accuracy", trControl = cvCtrl)
predicted_nb <- predict(bank_cv_nb, testSet, type = "raw")

results.matrix.nb <- confusionMatrix(predicted_nb, actual)


# nb drop education
head(testSet)
trainSetNB <- trainSet[-c(7)]
testSetNB <- testSet[-c(7)]
bank_cv_nb_reduced <- train(Personal.Loan ~ ., data = trainSetNB, method = "naive_bayes", metric = "Accuracy", trControl = cvCtrl)
predicted_nb_reduced <- predict(bank_cv_nb_reduced, testSetNB, type = "raw")

results.matrix.nb_reduced <- confusionMatrix(predicted_nb_reduced, actual)


#knn
bank_cv_knn <- train(Personal.Loan ~ ., data = trainSet, method = "knn", metric = "Accuracy", trControl = cvCtrl)
predicted_knn <- predict(bank_cv_knn, testSet, type = "raw")

results.matrix.knn <- confusionMatrix(predicted_knn, actual)


#dt
bank_cv_dt <- train(Personal.Loan ~ ., data = trainSet, method = "rpart", metric = "Accuracy", trControl = cvCtrl)
predicted_dt <- predict(bank_cv_dt, testSet, type = "raw")

results.matrix.dt <- confusionMatrix(predicted_dt, actual) 



print(results.matrix.nb)
print(results.matrix.knn)
print(results.matrix.dt)

print(results.matrix.nb_reduced)
```